---
title: "Analysis Report 2: The Relationships Among Possible Predictors and Causalities of Lung Cancer and its Different Stages"
author: "Charlotte Lopp"
date: "November 10, 2017"
output: github_document
bibliography: references.bib
csl: bioinformatics.csl
---

# Introduction

 In the 1950s, tobacco companies, with their ingenious advertisements promoting the idea that the quintessence of cool and glamour was or was at least very closely linked to cigarette smoking, started a trend that would change lives for decades. Hollywood icons, screen beauties, and even presidents were smoking and by the late 1950s around half of the population of industrialized nations smoked. The tobacco industry seemed untouchable and unstoppable but it all changed in 1994 when Diane Castano, whose husband died of lung cancer, sued the tobacco industry in the largest potential class action suit in history. Ever since, efforts have been made to expose not only the individual dangers of smoking but also the consequential dangers suffered by those who didn't decide to smoke - secondhand smoking, a concept championed by California politicians. Passive or indirect smoking may be one of the causal factors for lung cancer; it would explain the long-standing riddle of why many women (whose husbands were heavy smokers) develop lung cancer even though they themselves are non-smokers [@hirayama2000non]. 
  
Lung cancer is a leading cause of cancer-related death worldwide and one of the most common cancers in humans [@jemal2011global]. With the introduction of low-dose computerized tomography screening, early stage diagnosis is slowly increasing but lung cancer remains a devastating disease that has a very poor prognosis [@national2011reduced]. Non-small cell lung cancer (NSCLC) is the most common type of lunch cancer [@navada2006temporal] and accounts for approximately 85% of lung cancer cases in the USA [@sher2008small]. Adenocarcinoma is currently the most common type of NSCLC in Asian and North American patients and accounts for 30% of primary lung cancer in male smokers and 40% in female smokers in the USA [@seo2012transcriptional]. Smoking is the most prevalent risk factor for lung cancer, and the risk increases with the length of time and number of cigarettes people smoked [@powell2013association]. Research within the last decade has revealed differences in molecular characteristics of lung cancers in smokers versus nonsmokers, suggesting distinct pathogenic mechanism [@sun2007lung]. Cheng et al. found that overall gene expression levels were lower in smoker with lung cancer compared with healthy smokers in general [@cheng2012comparison]; Beane et al. found around 200 genes differentially expressed between eight smokers with lung cancer and five smoking healthy controls [@beane2011characterizing]; Han et al. identified 1063 genes that wee differentially expressed between tumor and normal tissues in smoking patients including SPP1, TOP2A, and CRABP2, etc [@han2014rna]. Despite all of them being valuable in adding literature and candidate genes in the collaborative study of lung cancer, non of them focused on identifying differences in gene expression profile between smokers and nonsmokers with lung cancer. Its the Seo et al. study in 2012 that looked at smokers and nonsmokers and utilized a hierarchical clustering method to identify cancer-up, cancer-down, and mixed regulated genes for each patient with extremely overexpressed genes also detected [@seo2012transcriptional].  
  
RNA-seq analysis provides us with a revolutionary tool for transcriptome analysis and with a novel opportunity to address such inquiries such a differences in gene expression profiles between smoking and nonsmoking patients [@wang2009rna]. RNA-seq offers several key advantages over existing technologies: 1) unlike hybridization-based approaches, RNA-seq is not limited to detecting transcripts that correspond  to existing genomic sequence (which makes it particularly attractive for non-model organisms with genomic sequences that are yet to be determined); 2) RNA-seq has very low, if any, background signal because DNA sequences can be unambiguously mapped to unique regions of the genome; 3) the results of RNA-seq also show high levels of reproducibility, for both technical and biological replicates; and 4) because there are no cloning steps, and with the Helicos technology there is no amplication step, RNA-seq requires less RNA sample [@wang2009rna]. Thus, using this powerful tool, Li et al. performed a large scale transcriptome analysis to identify the genes with altered expression in lung cancer development.


# Methods

## Sample origin and sequencing

In 2012, Seo et al. conducted transcriptome analysis of lung adenocarcinoma in order to identify the somatic mutations and transcriptional variations associated with lung cancer [@seo2012transcriptional]. They collected 200 fresh surgical specimens of primary lung adenocarcinoma from Korean patients who underwent major lunch resection and recorded diagnosis, gender, cancer stage, and smoking status for each patient. Among these 200 cancer patients, 54.5% (n = 109) were females and 58.0% (n = 116) were never-smokers. Of these, 87 cancer tissues whose driver mutations were not detected by screening tests were analyzed by transcriptome sequencing combined with whole-exome (n = 76) and transcriptome (n = 77) seuqencing of matched normal lung tissue samples. Transcriptome sequencing was a powerful method to use for detecting driver mutations in cancer, since not only somatic point mutations but also aberrant RNA variants such as fusion genes and alternative splicing can be examined. All these sequencing experiments were done as described previously. They generated 14,038,673,860 paired-end 101-bp-long reads from RNA sequencing of 164 samples (87 cancer and 77 corresponding normal tissues). On average, the RNA sequencing throughputs were 9.77 and 7.38 Gbp for cancer and normal tissues, respectively. In the whole-exome sequencing of normal tissues, they obtained 32.96-fold read depth per tissue for regions targeted by the exome capture platform used in this study. 
  
Li et al. [@li2015rna] downloaded Seo et al.'s RNA-seq data from paired normal and tumor tissues from 34 nonsmoking and 34 smoking patients with lung adenocarcinoma (GEO:GSE40419). They aimed to perform a large scale transcriptome analysis and identify the genes with altered expression in lung cancer development. They also looked for differences in gene expression profiles between smoking and nonsmoking patients in order to better define the impact of smoking in lung cancer development and provide insights that might better individualize treatment strategies for the patients.  


## Computational

After obtaining the RNA-seq data, Li et al. applied a stringent filter on the data to remove the gene tags with sparse count data. They then used edgeR to identify the genes differently expressed between tumor and normal tissues as well as the genes that varied between nonsmoker and smoker patients. The GLM model in edgeR allowed them to identify the genes that varied between nonsmoker and smoker patients with lung adenocarcinmoa. 
  
For our use of the data, our first step was to install the necessary tools and download sequencing files for RNA-Seq analysis. Next, we used R Script to install the biomartr package (from Bioconductor) and then used it to download a reference genome, transcripts, and annotations for the human genome from NCBI RefSeq. Biomartr [@drost2017biomartr] was used in this step because it implements straightforward functions for bulk retrieval of all genomci data for the selected files present in databases hosted by the National Center for Biotechnology Information (NCBI) and European Bioinformatics Institute (EMBL-EBI). Then a bash script was used to convert paired-end sra files to fastq files. The next step used sailfish; we created an index of kmers for sailfish quasi-aligner. Sailfish [@patro2014sailfish], which is a computational methd for quantifying the abundance of previously annotated RNA isoforms from RNA-seq data, was used because it provides quantification estimates much faster than other existing approaches without loss of accuracy. It exemplifies the potential of lightweight algorithms for efficiently processing sequencing reads by facilitating frequent reanalysis of data and reduicng the need to optimize parameters. We then ran fastqc on all fastq files and saved the output to an output directory using 86 threads. Trimmomatic was used to trim paired end reads in parallel. Trimmomatic [@bolger2014trimmomatic] is a program designed to handle paired-end data in a more flexible and efficient preprocessing way. Next was a script to run sailfish to count all reads from trimmed reads where both paired reads (F and R) made it through Trimmomatic QC. Then we needed to construct a transcript to gene ID mapping table. First was construction of a non-redundant list of Genbank ID to gene names. Second was to use grep over the first column in quant.sf files to find Genbank ID and then to add the gene name as second column. All of this gets loaded into R. 
  
The purpose of this R script is to read in all of the transcript counts made by sailfish individually on each sample, aggregate those counts (which are at the transcript level) to the gene level, normalize them based on length and read depth, and then build a table with rows as genes and columns as samples, with each cell being the normalized count for that gene in that sample. Then, we wanted to join in the two other metadata files we had, from SRA and from the supplementatry table from the original manuscript that has information like patient age and smoking behavior. We had to do some
cleaning along the way, and the gathered (melted) output file was rather big so we only included interesting columns, and wrote out the output in a compressed csv format as well as binary RData format.


# Results

```{r load-libraries, message = FALSE, echo = FALSE}
# Be sure to install these packages before running this script
# They can be installed either with the intall.packages() function
# or with the 'Packages' pane in RStudio

# load general-use packages
library("dplyr")
library("tidyr")
library("knitr")
library("ggplot2")
library("magrittr")

# this package allows for the easy inclusion of literature citations in our Rmd
# more info here: https://github.com/crsh/citr
# and here:
# http://rmarkdown.rstudio.com/authoring_bibliographies_and_citations.html
library("citr")
```

```{r load-data, message = FALSE, echo = FALSE}
# load the dataset from a compressed binary file
# it gets loaded as an object called "final_table"
# this has 3.4 million rows...so you will need to be thoughtful about
# how you analyze the data so that you don't overwhelm your laptop
load("output/final_compiled_counts/joined_count_data.RData")

# changes all columns names to lower case so Travis doesn't complain
names(final_table) %<>% tolower

# test that it loaded correctly before proceeding
stopifnot(exists("final_table"))
```

```{r make-summary-table, echo = FALSE}
# There are many many genes in this dataset, so we can
# subset it down to just a few here to look for interesting patterns
# in the most highly expressed
top_15 <- final_table %>%
  group_by(gender, genename) %>%
  summarize(mean_count = mean(counts_lengthscaledtpm)) %>%
  arrange(desc(mean_count)) %>%
  head(n = 15)

# then we can use the `kable()` function to make a nicely formatted
# markdown table
top_15 %>%
  kable()
```

**Figure 1**: This table is of the means of the most highly expressed genes in this dataset. 

```{r make-barplot-of-highly-expressed-genes, echo = FALSE}
# this code uses the same data as above, but use it to make a
# barplot - remember geom_col() is just like
# when you use geom_bar(stat = "identity")
top_15 %>%
  ggplot(aes(x = genename,
             y = mean_count,
             fill = gender)) +
    geom_col(position = "dodge") +
    xlab("Gene Name") +
    ylab("Mean Read Counts") +
    ggtitle("Mean Read Counts Per Gene By Gender") +
    theme(axis.text.x =
            element_text(angle = 90,
                         hjust = 1))
```

**Figure 2**: Here we see mean read counts per gene by gender; its a visualization of the data presented in Figure 1.  

```{r mean-and-sd-of-read-count-by-cancer-stage, echo = FALSE}
# this code looks at the mean and standard deviation of read count by cancer stage
final_table %>%
  group_by(cancer_stage) %>%
  summarize(mean_read_count = mean(counts_lengthscaledtpm),
            sd_read_count = sd(counts_lengthscaledtpm)) %>%
  ggplot(aes(x = cancer_stage,
             y = mean_read_count,
             color = cancer_stage)) +
      geom_errorbar(aes(ymin = mean_read_count - sd_read_count,
                        ymax = mean_read_count + sd_read_count),
                    width = 0.25) +
      geom_point(aes(y = mean_read_count)) +
      ylab("Mean Read Count") +
      xlab("Cancer Stage") +
      ggtitle("Mean and Standard Deviation of Read Count by Cancer Stage")
```

**Figure 3**: This figure shows the mean and standard deviation of the scaled read counts for each stage of cancer from 1A to 4.  

```{r mean-and-sd-of-read-count-by-age, echo = FALSE}
final_table %>%
  group_by(age_at_diagnosis) %>%
  summarize(mean_read_count = mean(counts_lengthscaledtpm),
            sd_read_count = sd(counts_lengthscaledtpm)) %>%
  ggplot(aes(x = age_at_diagnosis,
             y = mean_read_count,
             color = age_at_diagnosis)) +
      geom_errorbar(aes(ymin = mean_read_count - sd_read_count,
                        ymax = mean_read_count + sd_read_count),
                    width = 0.25) +
      geom_point(aes(y = mean_read_count)) +
      ylab("Mean Read Count") +
      xlab("Age at Diagnosis") +
      ggtitle("Mean and Standard Deviation of Read Count by Age at Diagnosis")
```

**Figure 4**: This figure shows the mean and standard deviation of the scaled read counts for age of diagnosis from 1A to 4. 

```{r count-by-age, echo = FALSE}
final_table %>%
  filter(genename == "A1BG") %>% 
  group_by(age_at_diagnosis) %>%
  tally() %>%
  ggplot(aes(x = age_at_diagnosis,
             y = n,
             fill = age_at_diagnosis)) +
    geom_col(position = "dodge") +
    xlab("Age at Diagnosis") +
    ylab("Count") +
    ggtitle("Distribution of Patients Across Age at Diagnosis")
```

**Figure 5**: This figure shows the distribution of patients across the vector of age at diagnosis.  

```{r make-facet-grid-of-age-by-length, echo = FALSE}
# this code looks at realtionship between cancer stage, read count, and age at diagnosis
final_table %>%
  ggplot(aes(x = age_at_diagnosis,
             y = counts_lengthscaledtpm)) +
      geom_point(size = 1, aes(color = gender)) +
      facet_grid(. ~ cancer_stage) +
      xlab("Age at Diagnosis") +
      ylab("Scaled Read Counts") +
      ggtitle("Read Counts by Age at Diagnosis and Cancer Stage by Gender") +
      theme(axis.text.x =
            element_text(angle = 90,
                         hjust = 1))
```

**Figure 6**: Here we show the relationship between scaled read counts and cancer stage, separated by age of diagnosis and further discriminated by gender.   

```{r age-by-cancer-stage-by-smoking-status, echo = FALSE}
# this code looks at the relationship among cancer stage, age of diagnosis and smoking status
final_table %>%
  ggplot(aes(x = cancer_stage,
             y = age_at_diagnosis)) +
      geom_point(size = 1, aes(color = smoking_status)) +
      ylab("Age at Diagnosis") +
      xlab("Cancer Stage") +
      ggtitle("The Relationship Between Age at Diagnosis and Cancer Stage by 
              Smoking Status")
```

**Figure 7**: This figure shows the relationship between stage of cancer and age at diagnosis, further separated by smoking status. 

```{r count-by-smoking-status, echo = FALSE}
final_table %>%
  filter(genename == "A1BG") %>% 
  group_by(smoking_status) %>%
  tally() %>%
  ggplot(aes(x = smoking_status,
             y = n,
             fill = smoking_status)) +
    geom_col(position = "dodge") +
    xlab("Smoking Status") +
    ylab("Count") +
    ggtitle("Distribution of Patients Across Smoking Status")
```

**Figure 8**: This figure shows the distribution of patients across the vector of smoking status.

```{r smoking-status-by-gender, echo = FALSE}
# this code looks at the relationship between gender and smoking status
final_table %>%
  filter(genename == "A1BG") %>% 
  group_by(smoking_status, gender) %>%
  tally() %>%
  ggplot(aes(x = smoking_status,
             y = n,
             fill = gender)) +
    geom_col(position = "dodge") +
    xlab("Smoking Status") +
    ylab("Count") +
    ggtitle("Relationship Between Gender and Smoking Status")
```

**Figure 9**: Here we see the relationship between gender and smoking status.  

```{r make-boxplot-of-highly-expressed-genes, echo = FALSE}
# here we just want to pull out the unique gene names and turn
# them into a vector so we can use it below to make a boxplot
# we use the pull() funtion to get this as a vector, just like
# we did when making histograms several weeks ago
top_genes <- top_15 %>%
  ungroup() %>%
  select(genename) %>%
  unique() %>%
  pull()

# now we need to filter from the full data set again, because
# we don't just want summary data, we want all the data in
# order to make boxplots
final_table %>%
  filter(genename %in% top_genes) %>%
  ggplot(aes(x = genename,
             y = counts_lengthscaledtpm,
             fill = gender)) +
    geom_boxplot() +
    facet_wrap(~cancer_stage) +
    xlab("Gene Name") +
    ylab("Scaled read counts per gene") +
    ggtitle("Read Counts Per Gene by Gender and Cancer Stage") +
    theme_bw() + # simplifies theme
    theme(axis.text.x = # rotates x axis labels vertically
            element_text(angle = 90,
                         hjust = 1))
```

**Figure 10**: This figure shows the relationship between scaled read counts per gene separated by cancer stage and further discriminated by gender.

## Summary of Results
The dataset is a table that has 3,434,400 points for 12 variables so it is a lot to be dealing with. Thus, our first step was to narrow down what we were looking out by separating the genes that were most highly expressed. Figure 1 is a table that details the top 15 most highly expressed genes along with which gender they were seen expressed in and their mean scaled read count arranged in descending order. The number one, most highly expressed gene is *EEF1A1*. 
  
We next compared mean read counts per gene by gender (Figure 2). The most highly expressed gene in both genders is *EEF1A1* followed by *SFTPB*. The rest are have noticeably lower expression. For the genes FN1, FTH1, and FTL, the expression is only seen in males and not females. For each gene, except SFTPA2 and SFTPB, the level of expression is higher in males than females. 
  
The next step was to looking at the means and stardard deviations of the scaled read counts for each stage of cancer from 1A to 4 (Figure 3). The means are all about the same with a slight increase in value for cancer stages 2B and 3A. These two stages also have the largest standard deviation (2B with the largest followed by 3A). The rest of the cancer stages have standard deviations that are significantly smaller - most of them fall between -2500 and 2500 (1B raches just above 2500). Their size in terms of standard deviation ranks in the following descending order: 2B, 3A, 1B, 1A and 2A are very similar, 2, and then 3B.

The next thing we wanted to look at was the relationship between these mean read counts and age at diagnosis (Figure 4). Here we see that that means are for the most part very similar to one another but there are discernable increases in the means for ages 45, 69, 70 and 78. The highest standard deviations are in ages 45, 65 and 69. If we looked into the distribution of patients across the vector of age at diagnosis (Figure 5), we see that the highest concentrations of patients were seen, in descending order, in the ages of 66, 62 and 58 with the number of patients respectively: 8, 7, 6. The next highest concentration was held by the three ages that had all the same number of patients of 5: 65, 68, 73.
  
We then wanted to look at the relationship between scaled read counts and cancer stage, separated by age of diagnosis and further discriminated by gender (Figure 6). In the cancer stage 2B, there seems to be only males. 3A has the highest read counts of all the stages and these higher read counts are all male; also, there is a clear deliniation between genders with females falling between the ages of 50 and 60 (in terms of age of diagnosis) and men between 60 and 80. In 1A, 1B and 2A, there is more of a mixture and overlap in terms of age of diagnosis. 2A, 3B and 4 all fall beneath read counts of 2.5e+05. Most of the read counts fall below 5e+05. For those with read counts higher than 5e+05, almost all of them are males except for a few females which are seen in cancer stage 1B. Those whose cancer stage is unknown have the least variance in terms of age: they are all around the age of 60. 1A, 2A, 3A, 4 and NA show the trend that the youngest people in each of these cancer stages are females; 1B somewhat shows this but it shows some that are younger being males. Both 2B and 3B show the youngest groups being males although 2B is made up entirely of males. In 1B, 2A, 2B, 3A, 4, and NA, the oldest people are males; only in 1A are the oldest females.
  
We were curious about the relationship between stage of cancer and age at diagnosis, further separated by smoking status (Figure 7). Here we see that anyone over 80 regardless of cancer stage is of the smoking status "previously." 3B only has people who have never smoked. 2B consists of "current" or "previously." Those who have never smoked are seen in each stage except for 2B from just below the age of 40 up to age 75. Those currently smoking are seen from 1A to 3A between the ages of just below 40 to just below 75. Those who have previously smoked are also seen in all age groups (and dominate the upper age groups) and are presented in each stage except 3B. Those whose smoking status is unknown are only seen in the unknown cancer stage and 3A between the ages of 60 and 70. It would seem that the overall concentration of lies within those who have either never smoked or previously smoked between the ages of 60 and 70. In order to look at the relative distribution of patients between smoking statuses, we look at Figure 8. Here we see that just below 40 have never smoked, just above 35 have smoked previously, just above 15 are currently smoking and just below 5 are unknown. 
  
In accordance with the trend mentioned in the introduction of women who had never smoked developing lung cancer, we then looked ats the relationship between gender and smoking status (Figure 9). For unknown smoking status, only males are represented. For those who have previously smoked, almost all are male - just over 30 are male and about 2 or 3 are females. Around the same trend is seen for those who have never smoked but inversed; the majority is female with just over 30 being female and just under 10 being male. For those currently smoking, about 1 is female and 15 are male.
  
The last thing that we looked at was the relationship between scaled read counts per gene separated by cancer stage and further discriminated by gender (Figure 10). The data represented in this figure starts with the top 15 genes, from the final table, which were enumerated in Figure 1. From this table, we pulled the unique gene names and turned them into a vector which were used as our x axis values. For cancer stages of 2A, 3B, 4, and NA, you can see that very low with counts all falling beneath 250,000. 3A shows some increase with counts reaching up to 500,000. 2B and 4 seem to have boxplots that are only male; 2A has boxplots that are only female. 1A is the next increment for higher count with the highest count belonging to gene EEF1A1 at 750,000. There is also one discernable boxplot for gene SFTPB that is male. Both 1B and 2B have counts that are above 750,000: for 1B it is seen in SFTPB and for 2B it is seen in SFTPA2. Overall, 2B has more longer counts but 1B is the only one with the largest boxplots, seen in genes SFTPB, that are both male and female.

# Discussion

Add around 1-2 pages interpreting your results and considering future directions one might take in analyzing these data.

Smoking is the most prevalent risk factor for lung cancer, and the risk increases with the length of time and number of cigarettes people smoked [@powell2013association]. Lung cancer, one of the most common cancers in humans, is a leading cause of cancer-related death worldwide [@jemal2011global]. Non-small cell lung cancer (NSCLC) is the most common type of lunch cancer [@navada2006temporal] and accounts for approximately 85% of lung cancer cases in the USA [@sher2008small]. Adenocarcinoma is currently the most common type of NSCLC in Asian and North American patients and accounts for 30% of primary lung cancer in male smokers and 40% in female smokers in the USA [@seo2012transcriptional]. Li et al. were looking for the genes with altered expression in lung cancer development as well as the differences in gene expression profiles between smoking and nonsmoking patients in order to better define the impact of smoking in lung cancer development and provide insight that might better individualize treatment strategies for the patients. 

The highest expression of genes in both males and females was seen in the genes EEF1A1 and SFTPB (Figure 2). For both of these genes, perhaps they are responsible for altered expression in lung cancer development with the foremost influential one being EE1A1 followed by SFTPB. For FN1, FTH1, and FTL, their expression was solely expressed in men. which means that their influence in altered expression is only seen in men. There were 9 most highly expressed genes overall and for all of them except two, they had higher mean read counts in men than women. This raises the question as to why. Is this because men have a higher probability of developing lung cancer? Or perhaps this is because are simply historically and culturally the more prevalent sex to pick up the habit of smoking. Perhaps it is the timeless quandry of the chicken or the egg. 

The continuation of our investigation takes us to calculating the mean and standard deviation of read count by cancer stage (Figure 3). The highest means are seen in both cancer stages 2B and 3A with the rest being signficantly smaller in their range. The holder of third place is 1B. This suggests that the genes with the highest rates of altered expression are seen in the greatest number (although this margin for means is small) in cancer stages 2B and 3A; their respective variations from this mean is largest in the two aforementioned cancer stages. Taking a look at Figure 10, we can see that this representation of the data makes sense. In 2B, we see large boxplots in comparison to the boxplots in the rest of the cancer stages with those for EEF1A1 and SFTPB having the largest boxplots. They are also wider because they are exclusively male. In 1B, you can see that for SFTPB the largest boxplot is female - again supporting that this gene has the greatest influence in altered expression in women. 

If we then look into how age factors into all of this, we see continued support for our previous findings. In Figure 6, we see that cancer stage 2B is completely male. 2B and 3A have the highest scaled read counts with 3A also being predominantly male. For almost all the stages, except for perhaps 3B and 1B do we see a male dominance. There is a concentration for age at diagnosis between the ages of 50 and 70. For more specificity and precision in terms of where the spikes are seen in terms of age at diagnosis, we look at Figure 4. Here we see that most of the means are the same but largest increases in mean read count are seen at the ages of 45, 69, 70 and 78 with the highest standard deviations seen in the ages of 45, 65 and 69. This seems to suggest that these ages have the higher altered expression and are therefore see a higher trend in development of lung cancer. If we looked into the distribution of patients across the vector of age at diagnosis (Figure 5), we see that the highest concentrations of patients were seen, in descending order, in the ages of 66, 62 and 58. The next highest concentration was held by the three ages that had all the same number of patients of 5: 65, 68, 73. This would seem to suggest that the 60s is the age range that sees the highest rate of lung cancer diagnoses. 

Now connecting age at diagosis, cancer stage, and smoking status, we turn to Figure 7. The trend that was established in Figures 4, 5 and 6 can be seen here in that the greatest concentration of patients were seen within the ages of 60 and 70. Here, you can also see that the smoking status of "never" is seen in every single cancer stage except 2B. The greatest number of non-smokers are seen in cancer stages 1A and 1B. Even though they have the greatest concentration in these early stages, non-smokers actually make up more than half of the patients that are seen in the later stages of 3B and 4. Those who have previously smoked are also present in every single stage (except 3B). We believe that the presence of these statuses in every stage except one each is a matter of sample size and that as the size of the study increased, the more evened out the trend would be of equal dispersal.  This dispersal of non-smokers and those that have previously smoked throughout all stages of cancer (except one) seems to suggest that to some extent the status of smoking does not matter. Granted, there are far more non-smokers with the beginning stages of lung cancer than any other stages but when looking at this data it is also important to consider the relative distribution between smoking status of the patients (Figure 8). Here we see that just below 40 have never smoked, just above 35 have smoked previously, just above 15 are currently smoking and just below 5 are unknown. Thus, of course we see more people that have smoked or have never smoked than those who are currently or those who are unknown. This leads us to question what kind of figure we would see if there was an equal distribution across all the smoking statuses. 

The results shown in Figure 9 confirm what was suggested, as was mentioned in the introduction, by previous literature - that there is indeed a high count of women who have never smoked that develop lung cancer. Passive or indirect smoking may be one of the causal factors for lung cancer; it would explain the long-standing riddle of why many women (whose husbands were heavy smokers) develop lung cancer even though they themselves are non-smokers [@hirayama2000non]. It raises the question of whether or not these women, the non-smokers that developed lung cancer, had partners or husbands that smoked or were in some way had consistent exposure to secondhand smoke. The same goes for the non-smoking men who developed lung cancer but as there's a higher percentage of men that smoke than women, this trend for the female sex in terms of societal impacts (like gender roles) as potential causation factors of lung cancer development due to second-hand smoke is more relevant. 

# Sources Cited
